---
title: "Introduction to Time-Frequency Decomposition"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
srate <- 2000
time <- seq(0, 4, by = 1/srate)
n <- length(time)
heatmap_pal <- colorRampPalette(c("#053061", "#2166ac", "#4393c3", "#92c5de", "#d1e5f0", 
"#ffffff", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f"
))(101)
spec_plot <- function(signal, srate = 2000, flim = 300, 
                      type = "h", main = "Signal", ...,
                      which = 1:3) {
  # Use another approach to calculate fft
  n <- length(signal)
  nyquist <- srate / 2
  n <- length(signal) # 2001
  frequency <- seq(0, nyquist, length.out = floor((n + 1) / 2))
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  
  
  # 2 rows of sub-figures, the second row has two figures
  layout(
    matrix(c(1,1,
             2,3), byrow = TRUE, nrow = 2)
  )
  par(mar = c(4.1, 4.1, 2.1, 0.1))
  
  # figure 1: raw trace of the signal
  if(1 %in% which) {
    plot(time, signal, type = "l", main = main, xlab = "Time (s)", ylab = bquote("Voltage ("~mu~"V)"), ...)
  }
  
  if(2 %in% which) {
    plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = type,
       main = "Frequency Domain")
  }
  
  if(3 %in% which) {
    plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = type, xlim = c(0, flim),
       main = sprintf("Zoom-in to 0.0~%.1f Hz", flim))
  }
}
tf_plot <- function(signal, srate = 2000, freq = seq(1,20, by = 0.5), cycle = c(1,4), main = "Time-Frequency Domain", log_power = TRUE, ...) {
  wavelet <- ravetools::morlet_wavelet(signal, freq, srate = srate, wave_num = cycle, precision = "float", trend = "linear")
  idx <- seq(1, length(signal), by = round(srate / 20))
  if(log_power) {
    amplitude <- 10*log10(Mod(wavelet[idx,,drop = FALSE]))
  } else {
    amplitude <- Mod(wavelet[idx,,drop = FALSE])
  }
  
  image(
    x = (idx - 1) / (round(srate / 20)),
    y = freq,
    z = amplitude,
    xlab = "Time (s)",
    ylab = "Frequency (Hz)",
    las = 1, main = main, ...
  )
}
repo <- raveio::prepare_subject_with_blocks('demo/DemoSubject', electrodes = 13:16, reference_name = "default", blocks = "008")
sample_signal <- repo$block_data$`008`$voltage$data[1:(2000 * 60), 2]



sample_frequency_2s <- seq(0, 1000, length.out = 2000)[1:400]
sample_time_2s <- seq(0, 58, by = 0.1)
sample_amplitude_2s <- t(sapply(sample_time_2s, function(start) {
  signal <- sample_signal
  srate <- 2000
  w <- (start + c(0, 2)) * srate
  w[1] <- max(w[1], 1)
  w[2] <- min(w[2], length(signal))
  signal <- signal[w[1] : w[2]]
  n <- length(signal)
  nyquist <- srate / 2
  n <- length(signal) # 2001
  frequency <- seq(0, nyquist, length.out = floor((n + 1) / 2))
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  fft_amplitude[1:400]
}))

sample_power_2s <- 20 * log10(sample_amplitude_2s)
```


## Recap

1. What's Fourier transform, DFT, FFT, Nyquist, amplitude
2. When should/shouldn't we use Fourier transform
3. What's Welch's method, why we should use Welch's method
4. What are two important parameters to set in Welch's method

## Why time-frequency decomposition?


```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 12,
    shiny::selectInput(
      inputId = "fft_signal_type",
      label = "Choose a signal",
      choices = c(
        "Louder",
        "Pulse",
        "Chirp",
        "Reverse Chirp",
        "Two states"
      )
    ),
    shiny::checkboxInput(
      inputId = "fft_include_noise",
      label = "Include noise"
    )
  ),
  
  shiny::column(
    width = 12,
    shiny::plotOutput("plot_fft_fail")
  )
)
```

```{r, context='server'}
output$plot_fft_fail <- shiny::renderPlot({
  set.seed(1)
  signal <- switch(
    input$fft_signal_type,
    "sin(2pi t) + cos(10 * 2pi t)" = {
      signal <- sin(2*pi * time) + cos(10 * 2*pi * time)
    },
    "sin(6 t) + cos(60 t)" = {
      signal <- sin(6 * time) + cos(60 * time)
    },
    "10 t - 15" = {
      signal <- 10*time - 15
    },
    "Pulse" = {
      signal <- sin(2*pi * time) + cos(10 * 2*pi * time)
      signal <- signal * dnorm(seq(-n/2, n/2-1)) * 30
      signal
    },
    "Chirp" = {
      sin(2 * pi * (2*time + time^2))
    },
    "Reverse Chirp" = {
      rev(sin(2 * pi * (2*time + time^2)))
    },
    "Louder" = {
      signal <- sin(2*pi * time) + cos(10 * 2*pi * time)
      signal * time*2 / 4
    },
    "Two states" = {
      s1 <- sin(2*pi * time)
      s2 <- sin(10 * 2*pi * time)
      sel <- (time > 1 & time < 2) | (time > 3 & time < 4)
      s1[sel] <- s2[sel]
      s1
    }
  )
  
  if(isTRUE(input$fft_include_noise)) {
    signal <- signal + rnorm(n)
  }
  
  spec_plot(signal, srate, 15, main = input$fft_signal_type, which = c(1,3))
  tf_plot(signal, srate, col = heatmap_pal[-(1:50)], log_power = FALSE)
})
```

How frequency components change over time?

## Welch's method (continued) 

```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 12,
    shiny::plotOutput(
      outputId = "pwelch_plot1", height = "250px"
    )
  ),
  shiny::column(
    width = 6,
    shiny::plotOutput(
      outputId = "pwelch_plot2", height = "250px"
    )
  ),
  shiny::column(
    width = 6,
    shiny::plotOutput(
      outputId = "pwelch_plot3", height = "250px"
    )
  ),
  shiny::column(
    width = 8,
    shiny::sliderInput(
      inputId = "pwelch_time_window_start",
      label = "Window starts (duration 2s)",
      min = 0, max = 58, step = 0.5,
      value = 0, animate = TRUE
    )
  ),
  shiny::column(
    width = 4,
    shiny::checkboxInput(
      inputId = "pwelch_baseline",
      label = "Baseline signal",
      value = FALSE
    )
  )
)
```

```{r, context='server'}
output$pwelch_plot1 <- shiny::renderPlot({
  ravetools::plot_signals(signals = sample_signal, sample_rate = 2000, space = 1)
  cex <- 1
  par(mar = c(3.1, 2.1, 2.1, 0.8) * (0.25 + cex * 0.75) + 0.1,
    mgp = cex * c(2, 0.5, 0))
  abline(v = input$pwelch_time_window_start + c(0, 2), col = 'red')
})


output$pwelch_plot2 <- shiny::renderPlot({
  start <- input$pwelch_time_window_start
  idx <- which.min(abs(start - sample_time_2s))
  fft_amplitude <- sample_amplitude_2s[idx, ]
  frequency <- sample_frequency_2s
  power <- 20 * log10(fft_amplitude)
  
  layout(matrix(c(1,2), nrow = 2), heights = c(1, lcm(3)))
  par(mar = c(0, 4.1, 5.1, 2.1))
  plot(x = frequency, 
       y = power,
       # xlab = "Frequency (Hz)", 
       ylab = "Power (dB)", type = 'l', ylim = c(-40, 40),
       main = "FFT on slice", axes = FALSE)
  axis(2, pretty(power), las = 1)
  par(mar = c(4.1, 4.1, 0, 2.1))
  power[power > 20] <- 20
  power[power < -20] <- -20
  image(x = frequency, z = matrix(power, ncol = 1), zlim = c(-20,20), col = heatmap_pal, axes = FALSE, ylab = "", xlab = "Frequency (Hz)")
  axis(1, pretty(frequency), las = 1)
})


output$pwelch_plot3 <- shiny::renderPlot({
  start <- input$pwelch_time_window_start
  idx <- which.min(abs(start - sample_time_2s))
  hit <- seq_len(idx)
  plot_data <- array(NA, dim(sample_power_2s))
  plot_data[hit, ] <- sample_power_2s[hit, ]

  # baseline
  if(isTRUE(input$pwelch_baseline)) {
    plot_data <- t((t(plot_data) - colMeans(sample_power_2s)) / apply(sample_power_2s, 2, sd))
    plot_data[plot_data > 4] <- 4
    plot_data[plot_data < -4] <- -4
    zlim <- c(-4,4)
  } else {
    plot_data[plot_data > 20] <- 20
    plot_data[plot_data < -20] <- -20
    zlim <- c(-20,20)
  }
  
  
  image(x = sample_time_2s, y = sample_frequency_2s,
        z = plot_data, zlim = zlim,
        axes = FALSE, xlab = "Time (s)", las = 1,
        ylab = "Frequency (Hz)", col = heatmap_pal)
  axis(1, pretty(sample_time_2s), las = 1)
  axis(2, pretty(sample_frequency_2s), las = 1)
})
```

## Time-Frequency: STFFT

```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 12,
    shiny::plotOutput(outputId = "stfft_plot0")
  )
)
```


```{r, context = "server"}
output$stfft_plot0 <- shiny::renderPlot({
  window <- input$stfft_window
  freq <- input$stfft_freq
  idx <- window * 2000; idx <- seq(idx[1], idx[2])
  time_center <- mean(window)
  sample_time <- seq(0, by = 1/2000, length.out = 20000)
  signal <- switch(
    input$stfft_signal,
    "spindle" = {
      s <- sin((sample_time - 7) * 2 * pi * 3) * dnorm(sample_time, mean = 7, sd = 0.3)
      space <- max(abs(s)) * 2
      s
    },
    {
      signal <- sample_signal[1:20000]
      space <- max(abs(signal)) * 2
      signal
    }
  )
  
  signal[-idx] <- 0
  sine_wave <- sin(2*pi*freq* (sample_time-time_center))
  cosine_wave <- cos(2*pi*freq* (sample_time-time_center))
  
  fft_coef <- (
    sum(cosine_wave * signal) +
      1i* sum(sine_wave * signal)
  ) / length(signal)
  
  cex <- 1
  par(cex = cex, cex.lab = 1,
      mar = c(3.1, 2.1, 2.1, 0.8) * (0.25 + cex * 0.75) + 0.1,
      mgp = cex * c(2, 0.5, 0), mfrow = c(3, 2))
  
  ravetools::plot_signals(signal, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(x(t)), space = space, space_mode = 'absolute')
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  plot(c(0,1), c(0,1), type = "n", axes = FALSE, xlab = "", ylab = "", main = "")
  text(x = 0.5, y = 0.8, sprintf("frequency=%.0fHz, time=%.1fs", freq, time_center))
  text(x = 0.5, y = 0.5, sprintf("coef=%.2f+%.2fi, amplitude=%.2f", Re(fft_coef), Im(fft_coef), Mod(fft_coef)))
  
  ravetools::plot_signals(cosine_wave, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(cos(2~pi~ft)~", "~f~"="~.(freq)))
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  ravetools::plot_signals(cosine_wave * signal, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(s(t)~".*"~sin(2~pi~ft)~", "~a[f]~.(sprintf("=%.2f", sum(cosine_wave * signal) / length(signal)))), space = space, space_mode = 'absolute')
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  ravetools::plot_signals(sine_wave, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(sin(2~pi~ft)~", "~f~"="~.(freq)))
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  ravetools::plot_signals(sine_wave * signal, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(s(t)~".*"~sin(2~pi~ft)~", "~b[f]~.(sprintf("=%.2f", sum(sine_wave * signal) / length(signal)))), space = space, space_mode = 'absolute')
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
})

```


> Instead of slicing input signals, we slice basis functions. Both yield the same results. 

Since we slice the basis functions, this approach is called "Short-time Fourier Transform" (STFFT)

```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 4,
    shiny::selectInput(
      inputId = "stfft_signal",
      label = "Input Signal",
      choices = c("sample", "spindle"),
      selected = "sample"
    )
  ),
  shiny::column(
    width = 4,
    shiny::sliderInput(
      inputId = "stfft_freq",
      label = "Frequency",
      min = 1, max = 20, step = 1,
      value = 1, animate = TRUE
    )
  ),
  shiny::column(
    width = 4,
    shiny::sliderInput(
      inputId = "stfft_window",
      label = "Window",
      min = 0, max = 10, step = 0.1,
      value = c(5, 7), animate = TRUE
    )
  ),
  shiny::column(
    width = 12,
    shiny::plotOutput(outputId = "stfft_plot1")
  )
)
```


```{r, context = "server"}
output$stfft_plot1 <- shiny::renderPlot({
  window <- input$stfft_window
  freq <- input$stfft_freq
  idx <- window * 2000; idx <- seq(idx[1], idx[2])
  time_center <- mean(window)
  sample_time <- seq(0, by = 1/2000, length.out = 20000)
  signal <- switch(
    input$stfft_signal,
    "spindle" = {
      s <- sin((sample_time - 7) * 2 * pi * 3) * dnorm(sample_time, mean = 7, sd = 0.3)
      space <- max(abs(s)) * 2
      s
    },
    {
      signal <- sample_signal[1:20000]
      space <- max(abs(signal)) * 2
      signal
    }
  )
  sine_wave <- sin(2*pi*freq* (sample_time-time_center))
  cosine_wave <- cos(2*pi*freq* (sample_time-time_center))
  sine_wave[-idx] <- 0
  cosine_wave[-idx] <- 0
  
  fft_coef <- (
    sum(cosine_wave * signal) +
      1i* sum(sine_wave * signal)
  ) / length(signal)
  
  cex <- 1
  par(cex = cex, cex.lab = 1,
      mar = c(3.1, 2.1, 2.1, 0.8) * (0.25 + cex * 0.75) + 0.1,
      mgp = cex * c(2, 0.5, 0), mfrow = c(3, 2))
  
  ravetools::plot_signals(signal, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(x(t)), space = space, space_mode = 'absolute')
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  plot(c(0,1), c(0,1), type = "n", axes = FALSE, xlab = "", ylab = "", main = "")
  text(x = 0.5, y = 0.8, sprintf("frequency=%.0fHz, time=%.1fs", freq, time_center))
  text(x = 0.5, y = 0.5, sprintf("coef=%.2f+%.2fi, amplitude=%.2f", Re(fft_coef), Im(fft_coef), Mod(fft_coef)))
  
  ravetools::plot_signals(cosine_wave, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(cos(2~pi~ft)~", "~f~"="~.(freq)))
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  ravetools::plot_signals(cosine_wave * signal, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(s(t)~".*"~sin(2~pi~ft)~", "~a[f]~.(sprintf("=%.2f", sum(cosine_wave * signal) / length(signal)))), space = space, space_mode = 'absolute')
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  ravetools::plot_signals(sine_wave, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(sin(2~pi~ft)~", "~f~"="~.(freq)))
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
  
  ravetools::plot_signals(sine_wave * signal, sample_rate = 2000, channel_names = "", xlab = "", ylab = "", main = bquote(s(t)~".*"~sin(2~pi~ft)~", "~b[f]~.(sprintf("=%.2f", sum(sine_wave * signal) / length(signal)))), space = space, space_mode = 'absolute')
  abline(v = c(window, time_center), col = c("red", "red", "gray60"), lty = c(1,1,3), lwd = c(3,3,1))
})

```


## Time-Frequency, that's it?

This is just a start! Take a look at time-frequency estimation below using the same input signal.

```{r, echo = FALSE, out.height="800px", fig.height=8}
wavelet <- ravetools::morlet_wavelet(sample_signal, freqs = seq(2, 200, 2), srate = 2000, wave_num = c(3,20), precision = "float", trend = "linear", signature = "tutorial-demosubject-e14-subset-0-60")
wavelet_power <- 20*log10(Mod(wavelet[seq(1, by = 200, length.out = 581), ]))

par(mfrow = c(2, 1))
image(x = sample_time_2s, y = sample_frequency_2s, sample_power_2s, col = heatmap_pal, xlab = "Time (s)", ylab = "Frequency (Hz)", main = "STFFT Power")
image(x = sample_time_2s, y = seq(2, 200, 2), wavelet_power, col = heatmap_pal, xlab = "Time (s)", ylab = "Frequency (Hz)", main = "Wavelet Power by RAVE")
```

<summary>
<details>
What's missing?
</details>

* Kernel taper
* Time-frequency uncertainty/trade-off

We will cover them in the next sessions.

</summary>

