---
title: "Limitations of Fourier Transform"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
srate <- 2000
time <- seq(0, 4, by = 1/srate)
n <- length(time)
spec_plot <- function(signal, srate = 2000, flim = 300, 
                      type = "h", main = "Signal", ...) {
  # Use another approach to calculate fft
  n <- length(signal)
  nyquist <- srate / 2
  n <- length(signal) # 2001
  frequency <- seq(0, nyquist, length.out = floor((n + 1) / 2))
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  
  
  # 2 rows of sub-figures, the second row has two figures
  layout(
    matrix(c(1,1,
             2,3), byrow = TRUE, nrow = 2)
  )
  par(mar = c(4.1, 4.1, 2.1, 0.1))
  
  # figure 1: raw trace of the signal
  plot(time, signal, type = "l", main = main, xlab = "Time (s)", ylab = bquote("Voltage ("~mu~"V)"), ...)
  
  plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = type,
       main = "Frequency Domain")
  
  plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = type, xlim = c(0, flim),
       main = sprintf("Zoom-in to 0.0~%.1f Hz", flim))
}
repo <- raveio::prepare_subject_with_blocks('demo/DemoSubject', electrodes = 13:16, reference_name = "default", blocks = "008")
sample_signal <- repo$block_data$`008`$voltage$data[1:(2000 * 60), 2]
```


## Recap: Fourier Transform

<details>
<summary>1. What are the Fourier transform, DFT, and FFT?</summary>

> `Fourier transform` decomposes time-dependent data (continuous functions of time) into additions of a series of $\cos(\cdot)$ and $\sin(\cdot)$ signals

1. $\cos(2\pi ft)$ and $\sin(2\pi ft)$ signals are simple: once frequency $f$ is given, the signal shapes can be easily derived at any time $t$
2. For each frequency $f$, there is a corresponding Fourier coefficient $a_{f} + i b_{f}$ (complex number), whose absolute number represents "the strength of signal at frequency $f$" (amplitude)

> `Discrete Fourier transform` (DFT) is Fourier **transform applied to discrete data signals** (data with finite number of time points, i.e. finite sample rates and finite durations)

> `Fast Fourier transform` (FFT) is the **algorithm** to perform discrete Fourier transform 

</details>

<details>
<summary>2. Why Fourier transform?</summary>

> We believe the brain signals are mixtures of simpler signals. Fourier transform makes it easy to view those components from the frequency domain.

</details>

<details>
<summary>3. How many distinct frequencies I can obtain from discrete Fourier transform? What's Nyquist?</summary>

> DFT output length is the same as the input data. That is, if the input signal `x` has `n` data points, then the output length will be `n` (i.e. `n` distinct frequencies)

> However, around half of `DFT` frequencies are useless due to aliasing introduced by finite sample rate. These frequencies are called "negative frequencies". **Nyquist** is the highest 
frequency that is meaningful to us (sampleRate/2).

</details>

This session will discuss the limitations of Fourier transform.

The following R function will be used through out the slides.

```{r, echo = TRUE, collapse=TRUE, results='hide'}
spec_plot <- function(signal, srate = 2000, flim = 300, 
                      type = "h", main = "Signal", ...) {
  # Use another approach to calculate fft
  n <- length(signal)
  nyquist <- srate / 2
  n <- length(signal) # 2001
  frequency <- seq(0, nyquist, length.out = floor((n + 1) / 2))
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  
  
  # 2 rows of sub-figures, the second row has two figures
  layout(
    matrix(c(1,1,
             2,3), byrow = TRUE, nrow = 2)
  )
  par(mar = c(4.1, 4.1, 2.1, 0.1))
  
  # figure 1: raw trace of the signal
  plot(time, signal, type = "l", main = main, xlab = "Time (s)", ylab = bquote("Voltage ("~mu~"V)"), ...)
  
  plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = type,
       main = "Frequency Domain")
  
  plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = type, xlim = c(0, flim),
       main = sprintf("Zoom-in to 0.0~%.1f Hz", flim))
}
```

## Limitations of FFT

```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 12,
    shiny::selectInput(
      inputId = "fft_signal_type",
      label = "Choose a signal",
      choices = c(
        "sin(2pi t) + cos(10 * 2pi t)",
        "sin(6 t) + cos(60 t)",
        "Louder",
        "10 t - 15",
        "Pulse",
        "Chirp",
        "Reverse Chirp",
        "Two states"
      )
    ),
    shiny::checkboxInput(
      inputId = "fft_include_noise",
      label = "Include noise"
    )
  ),
  
  shiny::column(
    width = 12,
    shiny::plotOutput("plot_fft_fail")
  )
)
```

```{r, context='server'}
output$plot_fft_fail <- shiny::renderPlot({
  set.seed(1)
  signal <- switch(
    input$fft_signal_type,
    "sin(2pi t) + cos(10 * 2pi t)" = {
      signal <- sin(2*pi * time) + cos(10 * 2*pi * time)
    },
    "sin(6 t) + cos(60 t)" = {
      signal <- sin(6 * time) + cos(60 * time)
    },
    "10 t - 15" = {
      signal <- 10*time - 15
    },
    "Pulse" = {
      signal <- sin(2*pi * time) + cos(10 * 2*pi * time)
      signal <- signal * dnorm(seq(-n/2, n/2-1)) * 30
      signal
    },
    "Chirp" = {
      sin(2 * pi * (2*time + time^2))
    },
    "Reverse Chirp" = {
      rev(sin(2 * pi * (2*time + time^2)))
    },
    "Louder" = {
      signal <- sin(2*pi * time) + cos(10 * 2*pi * time)
      signal * time*2 / 4
    },
    "Two states" = {
      s1 <- sin(2*pi * time)
      s2 <- sin(10 * 2*pi * time)
      sel <- (time > 1 & time < 2) | (time > 3 & time < 4)
      s1[sel] <- s2[sel]
      s1
    }
  )
  
  if(isTRUE(input$fft_include_noise)) {
    signal <- signal + rnorm(n)
  }
  
  spec_plot(signal, srate, 15, main = input$fft_signal_type)
})
```

* FFT works well when signals are "stationary" and finish "full cycle"
  - A standing wave, also known as a stationary wave, is a wave that oscillates in time but whose peak amplitude profile does not move in space. (Wikipedia)

## Welch's method


```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 12,
    shiny::plotOutput(
      outputId = "pwelch_plot1", height = "250px"
    )
  ),
  shiny::column(
    width = 6,
    shiny::plotOutput(
      outputId = "pwelch_plot2", height = "250px"
    )
  ),
  shiny::column(
    width = 6,
    shiny::plotOutput(
      outputId = "pwelch_plot3", height = "250px"
    )
  ),
  shiny::column(
    width = 12,
    shiny::sliderInput(
      inputId = "pwelch_time_window",
      label = "Time window",
      min = 0, max = 60, step = 0.5,
      value = c(0,2), animate = TRUE
    )
  )
)
```

```{r, context='server'}
output$pwelch_plot1 <- shiny::renderPlot({
  ravetools::plot_signals(signals = sample_signal, sample_rate = 2000, space = 1)
  cex <- 1
  par(mar = c(3.1, 2.1, 2.1, 0.8) * (0.25 + cex * 0.75) + 0.1,
    mgp = cex * c(2, 0.5, 0))
  abline(v = input$pwelch_time_window, col = 'red')
})

output$pwelch_plot2 <- shiny::renderPlot({
  signal <- sample_signal
  srate <- 2000
  n <- length(signal)
  nyquist <- srate / 2
  n <- length(signal) # 2001
  frequency <- seq(0, nyquist, length.out = floor((n + 1) / 2))
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  
  plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = 'l',
       main = "Direct FFT", xlim = c(0,80), ylim = c(0, 60))
})


output$pwelch_plot3 <- shiny::renderPlot({
  signal <- sample_signal
  srate <- 2000
  w <- input$pwelch_time_window * srate
  w[1] <- max(w[1], 1)
  w[2] <- min(w[2], length(signal))
  signal <- signal[w[1] : w[2]]
  n <- length(signal)
  nyquist <- srate / 2
  n <- length(signal) # 2001
  frequency <- seq(0, nyquist, length.out = floor((n + 1) / 2))
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  
  plot(frequency, fft_amplitude, xlab = "Frequency (Hz)", 
       ylab = "Amplitude", type = 'l',
       main = "FFT on slice", xlim = c(0,80), ylim = c(0, 60))
})
```

```{r, echo = FALSE}
sample_n <- length(sample_signal)
start <- seq(1, sample_n - 4000, by = 2000)
frequency <- seq(0, 1000, length.out = 2000)
amps <- sapply(start, function(i) {
  signal <- sample_signal[i -1 + 1:4000]
  srate <- 2000
  nyquist <- srate / 2
  
  fft_coef <- fft(signal) / length(signal)
  fft_coef <- fft_coef[seq_along(frequency)]
  
  fft_amplitude <- abs(fft_coef)
  fft_amplitude[-1] <- fft_amplitude[-1] * 2
  fft_amplitude
})

matplot(frequency, amps, xlim = c(0, 80), col = "gray60", lty = 1, type = 'l', main = "Welch's method", xlab = "Frequency (Hz)", ylab = "Amplitude", axes = FALSE)
lines(frequency, rowMeans(amps), col = 1)
axis(1, pretty(c(0, 80)))
axis(2, pretty(c(0, 150)))
```

### How to use RAVE to generate Welch's plot?

```{r, eval = FALSE}
library(ravetools)
pwel <- pwelch(sample_signal, fs = 2000, window = 4000, noverlap = 2000)
plot(pwel)
```

* This plot is generated on "log" scale:
  - Frequency is `log10` transformed
  - Power is in decibel `dB` unit: $10\log_{10}(\text{amp}^2)$

* The signal's Welch periodogram in log-scale is almost linear: you can fit a linear regression of power (dB) over log10 frequency - This is `1/f` curve.

## Play with parameters


```{r, echo = FALSE}
shiny::fluidRow(
  
  shiny::column(
    width = 12,
    shiny::plotOutput(
      outputId = "pwelch_demo1", height = "150px"
    )
  ),
  shiny::column(
    width = 12,
    shiny::plotOutput(
      outputId = "pwelch_demo2", height = "350px"
    )
  ),
  shiny::column(
    width = 6,
    shiny::sliderInput(
      inputId = "pwelch_window_size",
      label = "Pwelch window size",
      min = 0, max = 10, step = 0.5,
      value = 2, animate = TRUE, post = " s"
    )
  ),
  shiny::column(
    width = 6,
    shiny::sliderInput(
      inputId = "pwelch_noverlap",
      label = "Pwelch window overlap",
      min = 0, max = 90, step = 10,
      value = 50, animate = TRUE, post = "%"
    )
  )
)
```

```{r, context='server'}
output$pwelch_demo1 <- shiny::renderPlot({
  ravetools::plot_signals(signals = sample_signal, sample_rate = 2000, space = 1)
  srate <- 2000
  window <- round(max(input$pwelch_window_size * srate, 32))
  noverlap <- round(input$pwelch_noverlap * window / 100)
  step <- (window - noverlap) / srate
  cex <- 1
  par(mar = c(3.1, 2.1, 2.1, 0.8) * (0.25 + cex * 0.75) + 0.1,
    mgp = cex * c(2, 0.5, 0))
  abline(v = step * c(0:(60 / step)), lty = 1, col = "red")
  abline(v = step * c(0:(60 / step)) + (window / srate), lty = 3, col = "blue")
})

output$pwelch_demo2 <- shiny::renderPlot({
  srate <- 2000
  window <- round(max(input$pwelch_window_size * srate, 32))
  noverlap <- round(input$pwelch_noverlap * window / 100)

  pwel <- ravetools::pwelch(sample_signal, fs = srate, window = window, noverlap = noverlap, nfft = 2)
  plot(pwel, main = sprintf("Welch periodogram (w=%.1f sec,ovlp=%.1f sec)", window / srate, noverlap / srate), ylim = c(-40,40))
})
```
